Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-08-01 writeup/conStruct.tex
added poplar figs

Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, 
and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \citep{Frantz2009,Meirmans2012}.
In analyses of empirical datasets, which often are characterized by IBD,
model-based clustering approaches will therefore tend to overestimate the number of discrete clusters present,
with downstream implications for interpretation of results.
For example, conservation resources might be spread too thinly over areas 
that are spuriously assumed to be discrete management units.
Or, in studies of locally adaptive loci, or of the genetics of disease,
discrete populations may be modeled in separately,
reducing the global power to detect the genomic basis of complex phenotypes.

@@

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the nonspatial model 
run with between 1 and 7 clusters, applied to the dataset simulated under $K=1$,
and in Fig \ref{xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The admixture pie plots for the spatial and nonspatial models run on the datasets simulated under $K=2$ and $K=3$
are shown in Figs \ref{simK1_nsp_pies}-\ref{simK3_sp_pies}.

@@

In contrast, the predictive accuracy of the spatial model of \texttt{conStruct} 
either plateaus at or declines from the true value of $K$ used to simulate the data (Fig \ref{xvals}, second column).
And, as $K$ in the spatial model increases, 
deme membership in clusters beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
We parse \texttt{conStruct}'s performance on these simulation tests further in the Discussion.

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=8.5in]{figs/sims/xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The first panel in each row shows all results; 
	the second panel shows results for just the spatial model.
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{xvals}
\end{figure}

@@

>>>>

Existing model-based clustering approaches can only describe continuous patterns of variation using
discrete clusters, 
and so tend to erroneously describe continuously distributed variation with multiple clusters that 
show spatially autocorrelated cluster membership \citep{Frantz2009,meirmans2012}.
In analyses of empirical datasets, which often are characterized by IBD,
model-based clustering approaches will therefore tend to overestimate the number of discrete clusters present,
with downstream implications for interpretation of results.
For example, conservation resources might be spread too thinly over areas 
that are spuriously assumed to be discrete management units.
Or, in studies of locally adaptive loci, or of the genetics of disease,
discrete populations may be modeled in separately,
reducing the global power to detect the genomic basis of complex phenotypes.

@@

We ran \texttt{conStruct} analyses on each of these simulated datasets, 
specifying both a spatial and nonspatial model for $K = $ 1 through 7, 
then used cross-validation to compare the predictive performance of the models.
In Fig \ref{simK1_pies}, we show the results of the nonspatial model 
run with between 1 and 7 clusters, applied to the dataset simulated under $K=1$,
and in Fig \ref{sim_xvals} we show the results of the cross-validation analysis for each of the three simulated datasets.
The admixture pie plots for the spatial and nonspatial models run on the datasets simulated under $K=2$ and $K=3$
are shown in Figs \ref{simK1_nsp_pies}-\ref{simK3_sp_pies}.

@@

In contrast, the predictive accuracy of the spatial model of \texttt{conStruct} 
either plateaus at or declines from the true value of $K$ used to simulate the data (Fig \ref{sim_xvals}, second column).
And, as $K$ in the spatial model increases, 
deme membership in clusters beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
We parse \texttt{conStruct}'s performance on these simulation tests further in the Discussion.

@@

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=8.5in]{figs/sims/xvals.pdf}}
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The first panel in each row shows all results; 
	the second panel shows results for just the spatial model.
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    }\label{sim_xvals}
\end{figure}

@@

Trees in the genus \textit{Populus} (poplars, aspens, and cottonwoods) 
are distributed throughout the Northern Hemisphere, 
and include species of ecological and economic importance.
Species in the genus regularly co-occur and, 
where they do, they frequently hybridize (ref).
In addition, many \textit{Populus} species display patterns of isolation by distance, 
making the system well-suited to an analysis designed to tease apart 
genetic structure partitioned discretely across population (or species) clusters,
and continuously with geographic distance.

Here, we focus on two \textit{Populus} species in the Pacific Northwest: 
\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus trichocarpa}, the balsam poplar.
These species have a broad zone of overlap and are known to hybridize \citep{geraldes_etal_2014}.

We address this question using a dataset generated and originally published by \cite{geraldes_etal_2014}.
The published data consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera}).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data using both the spatial and nonspatial \texttt{conStruct} models, 
running on $K = 1$ through 7,
and compared these models using the cross-validation procedure described above. 
The results of these analyses are shown in 
Figs \ref{populus_xvals}-\ref{populus_nsp_clst_covs}. %,populus_nsp_pies,populus_sp_clst_covs,populus_nsp_clst_covs

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=5in]{figs/populus/populus_sampling_map.pdf}}
	\caption{
	Map of the sampled \textit{Populus} populations included in the analysis.
	The color of the sampling location denotes the putative species.
    }\label{populus_map}
\end{figure}

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=3in]{figs/populus/populus_std_xval.pdf}}
	\caption{
	Cross-validation results for \textit{Populus} dataset 
	comparing the spatial and nonspatial \texttt{conStruct} models run with $K=1$ through 7.  
	The first panel in each row shows all results; 
	the second panel zooms in on the results from analyses run with $K = 2$ through 7.
    }\label{populus_xvals}
\end{figure}

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{populus_sp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp2.pdf}}
		\subcaptionbox{$K=3$\label{populus_sp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp3.pdf}}
		\subcaptionbox{$K=4$\label{populus_sp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp4.pdf}}
		\subcaptionbox{$K=5$\label{populus_sp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp5.pdf}}
		\subcaptionbox{$K=6$\label{populus_sp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp6.pdf}}
		\subcaptionbox{$K=7$\label{populus_sp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_sp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{Populus} dataset 
	using the spatial model for $K=2$ through 7.
    }\label{populus_sp_pies}
\end{figure}

\begin{figure}
	\centering
		\subcaptionbox{$K=2$\label{populus_nsp_pies:K2}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp2.pdf}}
		\subcaptionbox{$K=3$\label{populus_nsp_pies:K3}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp3.pdf}}
		\subcaptionbox{$K=4$\label{populus_nsp_pies:K4}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp4.pdf}}
		\subcaptionbox{$K=5$\label{populus_nsp_pies:K5}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp5.pdf}}
		\subcaptionbox{$K=6$\label{populus_nsp_pies:K6}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp6.pdf}}
		\subcaptionbox{$K=7$\label{populus_nsp_pies:K7}}
			{\includegraphics[width=1.8in,height=1.8in]{figs/populus/populus_nsp7.pdf}}
	\caption{
	Map of admixture proportions estimated for the \textit{Populus} dataset 
	using the nonspatial model for $K=2$ through 7.
    }\label{populus_nsp_pies}
\end{figure}

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=4in]{figs/populus/populus_sp_clst_covs.pdf}}
	\caption{
	Plots showing the cluster-specific parametric covariance IBD-curves 
	estimated for the \textit{Populus} data using 
	the spatial \texttt{conStruct} model run with $K=2$ through 7.
	Colors are consistent with Fig \ref{populus_sp_pies}.
    }\label{populus_sp_clst_covs}
\end{figure}

\begin{figure}
	\centering
		{\includegraphics[width=6in,height=4in]{figs/populus/populus_nsp_clst_covs.pdf}}
	\caption{
	Plots showing the cluster-specific parametric covariance IBD-curves 
	estimated for the \textit{Populus} data using 
	the nonspatial \texttt{conStruct} model run with $K=2$ through 7.
	Colors are consistent with Fig \ref{populus_nsp_pies}.
    }\label{populus_nsp_clst_covs}
\end{figure}
